Reducing Linearizability of Priority Queues and More Data Structures to State Reachability

### 2017-04-18 co-regular-EPQ.tex
add files

A rule $R$ of extended priority queue is co-regular, if checking linearizability with respect to $\textit{MS}(R)$ of a data-independent implementation $\mathcal{I}$ can be reduced to checking the emptiness of intersection between $\mathcal{I}$ and a set of witness automata. In this section, we propose the definition of witness automata and co-regular. Then we prove that all five rules of extended priority queues are co-regular, and roughly introduce the proof idea. With the help of step-by-step linearizability and co-regular, we finally reduce the linearizability problem of $\textit{EPQ}$ into emptiness problem of intersection with automata.

\subsection{Definition of Co-Regular}
\label{subsec:definition of co-regular}

A witness automaton is a finite automaton with alphabet $\{ \textit{cal}(\textit{put},d,\textit{pred}), \textit{ret}(\textit{put},d),$ $\textit{cal}(\textit{rm},d),\textit{ret}(\textit{rm},d) \vert d \in \mathbb{D} \cup \{ \textit{empty} \},\textit{pred} \in \textit{predWA} \}$. Here $\textit{predWA}$ is the set of predicate of priorities, and it contains

\begin{itemize}
\setlength{\itemsep}{0.5pt}
\item[-] A predicate variable $p$, which accepts some specific priority,

\item[-] A predicate $\textit{les}_p$, which accepts all the priorities that are smaller than the value of $p$ according to $<_{\mathbb{P}}$,

\item[-] A predicate $\textit{anyPri}$, which accepts any priority.
\end{itemize}

Given a execution $e = \alpha_1 \cdots \alpha_k$ of extended priority queue and a witness automaton $\mathcal{A}$, we say that $e$ is accepted by $\mathcal{A}$, if

@@

\item[-] For each $i$, if $\alpha_i = \textit{cal}(\textit{put},a,q)$, then either (1) $\beta_i = \textit{cal}(\textit{put},a,p)$ and $q = d_p$, or (2) $\beta_i = \textit{cal}(\textit{put},a,\textit{les}_p)$ and $q <_{\mathbb{P}} d_p$, or (3) $beta_i = \textit{cal}(\textit{put},a,\textit{anyPri})$.

@@

Note that witness automata does not read operations, since the domains of operations is infinite.

Let us introduce the notion of co-regular:

%\vspace{-6pt}
\begin{definition}\label{def:co-regular of rules of extended priority queues}
A rule $R$ of extended priority queue is co-regular, if there are a finite set $\textit{Auts}_{R}$ of witness automata such that, for each data-independence implementation $\mathcal{I}$, we have that

$$\textit{Auts}_{R} \cap \mathcal{I} \neq \emptyset \Leftrightarrow \exists e \in \mathcal{I}_{\neq},e' \in \textit{proj}(e), R \in last(e') \wedge e' \ does \ not \ linearizable \ w.r.t. \ \textit{MS}(R)$$

We say that $\textit{EPQ}$ is co-regular, if each of its rule is co-regular.
\end{definition}

Before we go to investigate co-regular of each rules, we use the results in \cite{Bouajjani:2015} to simplify our work. \cite{Bouajjani:2015} states that checking linearizability w.r.t queue can be reduced into checking emptiness of intersection between $\mathcal{I}$ and a set of automata. Given a data-differentiated execution $e$, let $e \vert_{i}$ be an execution generated from $e$ by erasing call and return actions of items that does not use priority $i$ (does not influence $\textit{rm}(\textit{empty})$). We call a extended priority queue execution with only one priority a single-priority execution. Let $\textit{transToQueue}(e)$ be an execution generated from $e$ by transforming $\textit{put}$ and $\textit{rm}$ into $\textit{enq}$ and $\textit{deq}$, respectively, and then discarding priorities. We can see that for each $e \in \textit{EPQ}$ and each priority $i$, $\textit{transToQueue}(e \vert_{i})$ satisfy FIFO (first in first out) property.

Given an execution of queue, we say that it is differentiated \cite{Wolper:1986}, if each item is enqueued at most once. \cite{Bouajjani:2015} states that, given a differentiated queue execution $e$ without $\textit{deq}(\textit{empty})$, $e$ is not linearizable with respect to queue, if one of the following cases holds for some $a,b$: (1) $\textit{deq}(b) <_{hb} \textit{enq}(b)$, (2) there are no $\textit{enq}(b)$ and at least one $\textit{deq}(b)$, (3) there are one $\textit{enq}(b)$ and more than one $\textit{deq}(b)$, and (4) $\textit{enq}(a) <_{\textit{hb}} \textit{enq}(b)$, and $\textit{deq}(b) <_{\textit{hb}} \textit{deq}(a)$, or $\textit{deq}(a)$ does not exists. For each such case, we can construct a witness automata for extended priority queue. For example, for the first case, we generate witness automata $\mathcal{A}_{\textit{SinPri}}^1$ in \figurename~\ref{fig:automata for FIFO-1}, here $c_1 = \textit{cal}(\textit{put},a,\textit{anyPri}), \textit{ret}(\textit{put},a), \textit{cal}(\textit{rm},a)$, $\textit{ret}(\textit{rm},a),\textit{cal}(\textit{rm},b),\textit{cal}(\textit{rm},\textit{empty}),\textit{ret}(\textit{rm},\textit{empty})$, $c_2 = c_1 + \textit{ret}(\textit{rm},b)$, $c_3 = c_2 + \textit{ret}(\textit{put},b)$.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.6 \textwidth]{figures/PIC_AUTO_FIFO_1.pdf}
%\vspace{-10pt}
  \caption{Automaton $\mathcal{A}_{\textit{SinPri}}^1$}
  \label{fig:automata for FIFO-1}
\end{figure}

In Appendix \ref{sec:appendix proof and definition in section definition of co-regular}, we construct a set $\textit{Auts}_{\textit{sinPri}}$ of witness automata ($\mathcal{A}_{\textit{SinPri}}^1$ is in it), and shows that they are enough to ensure that for each data-differentiated executions, each of its single-priority projection without $\textit{rm}(\textit{empty})$ to have ``FIFO'' property, as shown by the following lemma.

@@

@@

In this subsection, we introduce the idea for proving co-regular of $\textit{PQ}_1^{>}$. The proof of this subsection can be found in Appendix \ref{sec:appendix proof and definition in section co-regular of EPQ1Lar}. The notion of left-right constraint used for extended priority queue is inspired by left-right constraint of queue \cite{Bouajjani:2015}.

Given a data-differentiated execution $e$, we say that $e$ is a $\textit{pri}$-execution, if $\textit{pri}$ is the maximal priority of $e$, and $\textit{pri}$ is larger than or equal to the priority of all other items of $e$. Given a data-differentiated execution $e$ and assume that $\textit{pri}$ is one of the maximal priority of $e$, let $\textit{pri-Exec}(e)$ be an execution obtained from $e$ by erasing all operations of items whose priority are incomparable with $\textit{pri}$. The following lemma states that when checking co-regular of $\textit{EPQ}_1^{>}$, $\textit{EPQ}_1^{=}$, $\textit{EPQ}_2^{>}$ and $\textit{EPQ}_2^{=}$, we need only consider $\textit{pri}$-executions.

Lemma \ref{lemma:pri execution is enough} simplifies our proof by make us safely ignore all the items that has priorities incomparable with $\textit{pri}$. When construct witness automata, priorities which are incomparable with $\textit{pri}$ can be safely represented by $\textit{anyPri}$. Given a $\textit{pri}$-execution $e'$, we can see that $\textit{last}(e')$ contains only one rules.

Given a data-differentiated $\_$-execution $e$ such that $\textit{last}(e) = \textit{EPQ}_1^{>}$, although Lemma \ref{lemma:automata for extended priority queue with single priority} ensures that each single-priority projection of $e$ satisfy the FIFO property, this is still not enough for ensuring that $e \sqsubseteq \textit{MS}(\textit{EPQ}_1^{>})$. Since it is possible that $e$ does not linearizable w.r.t $\textit{MS}(\textit{EPQ}_1^{>})$ because of interaction between actions of multiple priorities.

We give an example of such execution $e$ in \figurename~\ref{fig:introduce gap for EPQ1Lar}. We call the time interval from $\textit{ret}(\textit{put},x)$ to $\textit{cal}(rm,x)$, or from $\textit{ret}(\textit{put},x)$ when $\textit{cal}(rm,x)$ does not exist, the interval of item $x$. In \figurename~\ref{fig:introduce gap for EPQ1Lar}, we draw the interval of each item by dashed line. Here we assume that $p_1 <_{\mathbb{P}} p_4$, $p_2 <_{\mathbb{P}} p_4$ and $p_3 <_{\mathbb{P}} p_4$. The reason of why $e$ does not linearizable w.r.t $\textit{MS}(\textit{PQ}_1^{>})$ is that, each time point from $\textit{cal}(\textit{rm},b)$ to $\textit{ret}(\textit{rm},b)$ is in interval of some item with smaller priority.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.6 \textwidth]{figures/PIC-HIS-INTRO-GAP-EPQ1L.pdf}
%\vspace{-10pt}
  \caption{An execution that does not linearizable w.r.t $\textit{MS}(\textit{EPQ}_1^{>})$}
  \label{fig:introduce gap for EPQ1Lar}
\end{figure}

In this subsection, intuitively we will prove that, as long as we can get rid of the case in \figurename~\ref{fig:introduce gap for EPQ1Lar}, we can ensure $\textit{last}(e) = \textit{EPQ}_1^{>} \Rightarrow e \sqsubseteq \textit{MS}(\textit{EPQ}_1^{>})$.

Let us introduce the notion of left-right constraint, which is a graph, and the existence of cycle though item with maximal priority in it correspond to the existence of the case in \figurename~\ref{fig:introduce gap for EPQ1Lar}.

\begin{definition}\label{def:left-right constraint for matched put and rm operations}
Given a data-differentiated $\textit{pri}_x$-execution $e$ without $\textit{rm}(\textit{empty})$. Let $\textit{put}(x,\textit{pri}_x)$ and $\textit{rm}(x)$ be the only  method events of $e$ with priority $\textit{pri}_x$. The left-right constraint of $\textit{put}(x,\textit{pri}_x)$ and $\textit{rm}(x)$ is the graph $G$ where:

\begin{itemize}
\setlength{\itemsep}{0.5pt}
\item[-] the nodes are the items of $e$, to which we add a node,

\item[-] there is an edge from item $d_1$ to $x$, if $\textit{put}(d_1,\_)$ happens before $\textit{put}(x,\textit{pri}_x)$ or $\textit{rm}(x)$,

\item[-] there is an edge from $x$ to item $d_1$, if $\textit{rm}(x)$ happens before $\textit{rm}(d_1)$ or $\textit{rm}(d_1)$ does not exists in $h$,

\item[-] there is an edge from item $d_1$ to item $d_2$, if $\textit{put}(d_1,\_)$ happens before $\textit{rm}(d_2,\_)$.
\end{itemize}
\end{definition}

When there is a cycle $d_1 \rightarrow \ldots \rightarrow d_m \rightarrow x \rightarrow d_1$ through item with maximal priority (for example, $x$) in $G$, we say that $x$ is covered by $d_1,\ldots,d_m$. To state the effectiveness of left-right constraint, take the execution in \figurename~\ref{fig:introduce gap for EPQ1Lar} as an example, we can see that $b$ is covered by $f,e,d$. We need to prove that getting rid of cycle though item with maximal priority in left-right constraint is enough for ensure linearizable w.r.t $\textit{MS}(\textit{EPQ}_1^{>})$, as stated by the following lemma:

@@

The $\textit{only if}$ direction can be easily proved by contradiction and is omitted here. To prove the $\textit{if}$ direction, we need to explicitly construct the linearization of $e$, or we can say, we need construct the $u$, $v$ and $w$ in $\textit{EPQ}_1^{>}$. Let $u$ to be the sequence of all operations that happens before $\textit{put}(x)$. The difficulties is how to generate proper $v$.

To generate $v$, we introduce $\textit{UVSet}(e,x)$, which intuitively contains all pairs of method events that should be putted before $\textit{rm}(x)$. Let $\textit{UVSet}_1(e,x)= \{ o \vert$ either $o <_{\textit{hb}} \textit{put}(x)$ or $\textit{rm}(x)$, or $\exists o'$ with the same item of $o$, such that $o' <_{\textit{hb}} \textit{put}(x)$ or $\textit{rm}(x)\}$. For each $i \geq 1$, let $\textit{LMSet}_{\textit{i+1}}(e,x) = \{ o \vert$ $o \notin \textit{UVSet}_k(e,x)$ for each $k \leq i$, and either $o$ happens before some operation $o' \in \textit{UVSet}_i(e,x)$, or $\exists o''$ with the same item of $o$ and $o''$ happens before some operation $o' \in \textit{UVSet}_i(e,x)\}$. Let $\textit{UVSet}(e,x) = \textit{UVSet}_1(e,x) \cup \textit{UVSet}_2(e,x) \cup \ldots$. For example, assume that $p_1 <_{\mathbb{P}} p_2$, then in \figurename~\ref{fig:his nobound of UVSet}, $\textit{UVSet}_1(e,x) = \{ \textit{put}(a,p_1),\textit{rm}(a) \}$, $\textit{USSet}_2(e,x) = \{ \textit{put}(b,p_1),\textit{rm}(b) \}$ and $\textit{UVSet}_3(e,x) = \{ \textit{put}(c,p_1),\textit{rm}(c) \}$. Similarly, we can generate execution $e'$ such that, for each $i$, $\textit{UVSet}(e',x) \neq \emptyset$.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.5 \textwidth]{figures/PIC_HIS_NOBOUNDOF_UVSET.pdf}
%\vspace{-10pt}
  \caption{An example for $\textit{UVSet}(e,x)$}
  \label{fig:his nobound of UVSet}
\end{figure}

We let the $v$ to be the sequence of method events that are in $\textit{UVSet}(e,x)$ and not in $u$. We let $w$ to be the sequence of remaining method events. Then we can prove that such $u$, $v$ and $w$ holds as required. \qed
\end {proof}

According to Lemma \ref{lemma:Lin Equals Constraint for EPQ1Lar}, the existence of ``An item covered by items of smaller priority'' is enough for checking violation of linearizability w.r.t $\textit{MS}(\textit{EPQ}_1^{>})$. Assume $x$ is covered by $d_1,\ldots,d_m$. Then we can safely rename $x$ into $b$, rename $d_1,\ldots,d_m$ into $a$ and rename all other item into $c$ by data-independence. Such execution can be recognized by witness automata, since between the first $\textit{ret}(\textit{put},a,\_)$ and the last $\textit{cal}(\textit{rm},a)$ (if exists), $\textit{ret}(\textit{put},a,\_)$ and $\textit{cal}(\textit{rm},a)$ occurs in pair and there is no need for count.

There are four possible enumeration of call and return actions of $\textit{put}(b)$ and $\textit{rm}(b)$. For each of them, we generate a witness automaton. For example, for the case when $e \vert_{b} = \textit{cal}(\textit{put},b,p) \cdot \textit{ret}(\textit{put}) \cdot \textit{cal}(\textit{rm}) \cdot \textit{ret}(\textit{rm},b)$, we generate witness automaton $\mathcal{A}_{\textit{l-lar}}^1$, as shown in \figurename~\ref{fig:automata APQ1Lar-1}. Here $c_1 = c + \textit{ret}(\textit{rm},a)$, $c_2 = c + \textit{cal}(\textit{put},a,\textit{les}_p)$, $c_3 = c_2 + \textit{ret}(\textit{rm},a)$, where $c = \textit{cal}(\textit{put},d,\textit{anyPri}),\textit{ret}(\textit{put},d), \textit{cal}(\textit{rm},d), \textit{ret}(\textit{rm},d),\textit{cal}(\textit{rm},\textit{empty}), \textit{ret}(\textit{rm},\textit{empty})$.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=1 \textwidth]{figures/PIC_AUTO_PQ1Lar-pprr.pdf}
%\vspace{-10pt}
  \caption{Automaton $\mathcal{A}_{\textit{l-lar}}^1$}
  \label{fig:automata APQ1Lar-1}
\end{figure}

Given the execution $e$ in \figurename~\ref{fig:introduce gap for EPQ1Lar}, let $e'$ be generated from $e$ by renaming $d,e,f$ into $a$. Then we can see that $e'$ is accepted by $\mathcal{A}_{\textit{l-lar}}^1$ with a path via $q_{\textit{init}}, q_2, q_3, q_4,q_5,q_7,q_8$.

In Appendix \ref{sec:appendix proof and definition in section co-regular of EPQ1Lar}, we construct a set $\textit{Auts}_{\textit{1-lar}}$ of witness automata ($\mathcal{A}_{\textit{l-lar}}^1$ is in it), and use $\textit{Auts}_{\textit{1-lar}}$ to prove that $\textit{EPQ}_1^{>}$ is co-regular, as stated by the following lemma.

@@

In this subsection, we introduce the idea for proving co-regular of $\textit{EPQ}_1^{=}$. The proof of this subsection can be found in Appendix \ref{sec:appendix proof and definition in section co-regular of EPQ1Equal}.

Given a data-differentiated $\_$-execution $e$ such that $\textit{last}(e) = \textit{EPQ}_1^{=}$. Lemma \ref{lemma:automata for extended priority queue with single priority} ensures that each single-priority projection of $e$ satisfy the FIFO property, and with the help of Lemma \ref{lemma:EPQ1Lar is co-regular} and $\textit{Auts}_{\textit{1-lar}}$, we can ensure that each item in $e$ can not be covered by items with smaller priority. However, these are sill not enough for ensure that $e \sqsubseteq \textit{MS}(\textit{EPQ}_1^{=})$. This is because that given items $a$ and $b$ with maximal priority, it is possible that all the possible linearization point of $\textit{rm}(b)$ may be disabled by $\textit{rm}(b)$.

We give an example of such execution $e$ in \figurename~\ref{fig:introduce pb order}. The execution $e$ of \figurename~\ref{fig:introduce pb order} is not linearizable, even if $h \vert_{p_1}$ and $h \vert_{p_4}$ are both linearizable, and either $a$ or $b$ is covered by items with smaller priority. To explain this, me need to observe the following three points:

\begin{itemize}
\setlength{\itemsep}{0.5pt}
\item[-] Since $\textit{put}(a,p_4)$ happens before $\textit{put}(b,p_4)$, we know that $a$ is ``putted earlier'' than $b$, and therefore, the linearization points of $\textit{rm}(a)$ should before the linearization point of $\textit{rm}(b)$.

\item[-] The time intervals satisfy the following conditions (we identify them with dotted lines in \figurename~\ref{fig:introduce pb order}) are possible position to locate the linearization point of $\textit{rm}(b)$ according to Lemma \ref{lemma:Lin Equals Constraint for EPQ1Lar} (Here we temporarily forget the existence of $a$): (1) between $\textit{cal}(\textit{rm},b)$ and $\textit{ret}(\textit{rm},b)$, (2) does not in interval of any item with smaller priority.

\item[-] To satisfy requirement of the second condition, the linearization points of $\textit{rm}(b)$ should be in the time interval of dotted line. However, all time point in dotted line is before $\textit{cal}(\textit{rm},a)$ and is thus disabled by $\textit{cal}(\textit{rm},a)$.
\end{itemize}

@@

In this subsection, intuitively we will prove that, as long as we can get rid of the case in \figurename~\ref{fig:introduce pb order}, we can ensure $\textit{last}(e) = \textit{EPQ}_1^{=} \Rightarrow e \sqsubseteq \textit{MS}(\textit{EPQ}_1^{=})$.

Let us introduce $<_{\textit{pb}}$ (represents put-before) order to formally state that ``an item is putted before another item''. Given a data-differentiated $\_$-execution $e$ and two items $a,b$ with maximal priority in $e$, we say that $a <_{\textit{pb}} b$, if one of the following cases holds:

\begin{itemize}
\setlength{\itemsep}{0.5pt}
\item[-] $\textit{put}(a,\_)$ happens before $\textit{put}(b,\_)$ in $e$,

\item[-] $\textit{rm}(a)$ happens before $\textit{rm}(b)$ in $e$,

\item[-] $\textit{rm}(a)$ happens before $\textit{put}(b,\_)$ in $e$,
\end{itemize}

Sometimes we use $a <_{\textit{pb}}^A b$, $a <_{\textit{pb}}^B b$ and $a <_{\textit{pb}}^C b$ to explicitly distinguish above three cases. Let $<_{\textit{pb}}^*$ be the transitive closure of $<_{\textit{pb}}$. Intuitively, $a <_{\textit{pb}}^* b$ means that we can infer that $a$ should be inserted earlier than $b$ with the help of possibly other items.

@@

\begin{definition}\label{def:gap-point for matched put and rm operations}
Given a data-differentiated $\textit{pri}_x$-execution $e$ and two method events $\textit{put}(x,\textit{pri}_x),\textit{rm}(x)$ of $e$. We say that a time-point $o$ is a gap-point of $x$, if

\begin{itemize}
\setlength{\itemsep}{0.5pt}
\item[-] $o$ is after $\textit{call}(\textit{put},x,\textit{pri}_x)$ and $\textit{call}(\textit{rm},x)$, and is before $\textit{ret}(\textit{rm},x)$.

\item[-] $o$ is not in interval of any items with smaller priority.
\end{itemize}
\end{definition}

Going back to \figurename~\ref{fig:introduce pb order}, we can effectively characterize it using $<_{\textit{pb}}$ order and gap-point as follows: $a <_{\textit{pb}} b$, while the right-most gap-point of $b$ is before $\textit{cal}(\textit{rm},a)$. We need to prove that getting rid of such case is enough fo ensure linearizable w.r.t $\textit{MS}(\textit{PQ}_1^{=})$, as stated by the following lemma. Let $\textit{Items}(e,p)$ be the set of items with priority $p$ in execution $e$.

\begin{restatable}{lemma}{EPQOneEqualAsPBandGP}
\label{lemma:EPQ1Equal as pb order and gap-point}
Given a data-differentiated $\textit{pri}$-execution $e$ with $\textit{last}(e) = \textit{EPQ}_1^{=}$. $e$ does not linearizable w.r.t $\textit{MS}(\textit{EPQ}_1^{=})$, if and only if there exists $x$ and $y$ with maximal priority $\textit{pri}$ in $e$, such that $y <_{\textit{pb}}^* x$ in $e$, and the rightmost gap-point of $x$ is before $\textit{cal}(\textit{put},y,\textit{pri})$ or $\textit{cal}(\textit{rm},y)$ in $e$.
\end{restatable}

@@

We have already intuitively explain the proof of the $\textit{if}$ direction.

>>>>

In this section, we use the notation co-regular to reduce checking linearizability w.r.t $\textit{MS}(R)$ into checking the emptiness of intersection between implementation and a set of witness automata. With the help of step-by-step linearizability and co-regular, we finally reduce the linearizability problem of $\textit{EPQ}$ into state reachability problem.

\subsection{Witness Automata and Co-Regular}
\label{subsec:definition of co-regular}

$\textit{con-check-EPQ}$ inspires us that, for data-differentiated executions, checking linearizability with respect to $\textit{EPQ}$ is equivalent to checking linearizability with respect to $\textit{MS}(R)$ for everyone of its projections, as stated by the following lemma.

The set of projections of $e$ is denoted $\textit{proj}(e)$. When refer to $\textit{proj}(e)$, we implicitly assume that each $\textit{rm}(\textit{empty})$ in $e$ has a ghost argument that is unique.

\begin{restatable}{lemma}{EPQasMultiInMRpriforHistory}
\label{lemma:EPQ as multi in MRpri for history}
Given a data-differentiated execution $e$, $e \sqsubseteq \textit{EPQ}$, if and only if, $\forall e' \in \textit{proj}(e)$ and $\forall R \in \textit{last}(e')$, we have $e' \sqsubseteq \textit{MS}(R)$.
\end{restatable}

A witness automaton is a finite automaton with priority variable. It uses alphabet $\{ \textit{cal}(\textit{put},d,\textit{pred}), \textit{ret}(\textit{put},d),$ $\textit{cal}(\textit{rm},d),\textit{ret}(\textit{rm},d) \vert d \in \mathbb{D} \cup \{ \textit{empty} \},\textit{pred} \in \textit{predWA} \}$. Here $\textit{predWA} = \{p, \textit{les}_p, \textit{anyPri}\}$, where $p$ is a predicate variable, $\textit{les}_p$ represents priorities that are smaller than the value of $p$, and $\textit{anyPri}$ represents any priority. Given an execution $e = \alpha_1 \cdots \alpha_k$ of extended priority queue and a witness automaton $\mathcal{A}$, we say that $e$ is accepted by $\mathcal{A}$, if

@@

\item[-] For each $i$, if $\alpha_i = \textit{cal}(\textit{put},a,q)$, then either (1) $q = d_p$ and $\beta_i = \textit{cal}(\textit{put},a,p)$, or (2) $q <_{\mathbb{P}} d_p$ and $\beta_i = \textit{cal}(\textit{put},a,\textit{les}_p)$, or (3) $\beta_i = \textit{cal}(\textit{put},a,\textit{anyPri})$.

@@

Note that due to data-independence, witness automata are used to detect some renamed violations, instead of store and checking directly. Let us introduce the notion of co-regular, which reduce checking linearizable w.r.t $\textit{MS}(R)$ into checking set of witness automata.

%\vspace{-6pt}
\begin{definition}\label{def:co-regular of rules of extended priority queues}
Given $R \in \{ \textit{EPQ}_1^{>}, \textit{EPQ}_1^{=}, \textit{EPQ}_2^{>}, \textit{EPQ}_2^{=}, \textit{EPQ}_3 \}$. $R$ is co-regular, if there are a finite set $\textit{Auts}_{R}$ of witness automata such that, for each data-independence implementation $\mathcal{I}$, we have that

$$\textit{Auts}_{R} \cap \mathcal{I} \neq \emptyset \Leftrightarrow \exists e \in \mathcal{I}_{\neq},e' \in \textit{proj}(e), R \in last(e') \wedge e' \not\sqsubseteq \textit{MS}(R)$$

We say that $\textit{EPQ}$ is co-regular, if all five candidates of $R$ are co-regular.
\end{definition}

We simply the proof of co-regular of $\textit{EPQ}$ in two side. The first side uses the results in \cite{Bouajjani:2015} to ensure FIFO (first in first out) property of each single-priority projections. A execution where only items of one priority is putted is called a single-priority execution. It is obvious that each single-priority projection of sequences in $\textit{EPQ}$ satisfy FIFO property. Let $\textit{transToQueue}(e)$ be an execution generated from $e$ by transforming $\textit{put}$ and $\textit{rm}$ into $\textit{enq}$ and $\textit{deq}$, respectively, and then discarding priorities. In Appendix \ref{sec:appendix proof and definition in section definition of co-regular}, according to the result of queue in \cite{Bouajjani:2015}, we construct a set $\textit{Auts}_{\textit{sinPri}}$ of witness automata, and shows that they are enough to ensure that for each data-differentiated executions, each of its single-priority projection without $\textit{rm}(\textit{empty})$ to have ``FIFO'' property, as shown by the following lemma.

@@

The second side is to ignore the incomparable priorities. Given a data-differentiated execution $e$, we say that $e$ is a $\textit{pri}$-execution, if $\textit{pri}$ is the maximal priority of $e$, and $\textit{pri}$ is larger than or equal to the priority of all other items of $e$. Given a data-differentiated execution $e$ and assume that $\textit{pri}$ is one of the maximal priority of $e$, let $\textit{pri-Exec}(e)$ be an execution obtained from $e$ by erasing all operations of items whose priority are incomparable with $\textit{pri}$. The following lemma states that when checking co-regular of $\textit{EPQ}$, it is safe to consider $\textit{pri}$-executions. We can see that $\textit{last}(e')$ contains at most one element for $\textit{pri}$-execution $e'$.

@@

In this subsection, we briefly introduce the idea for proving co-regular of $\textit{PQ}_1^{>}$. The proof of this subsection can be found in Appendix \ref{sec:appendix proof and definition in section co-regular of EPQ1Lar}. The notion of left-right constraint used for extended priority queue is inspired by left-right constraint of queue \cite{Bouajjani:2015}.

Given a data-differentiated $\_$-execution $e$ such that $\textit{last}(e) = \textit{EPQ}_1^{>}$, Lemma \ref{lemma:automata for extended priority queue with single priority} is not enough for ensuring that $e \sqsubseteq \textit{MS}(\textit{EPQ}_1^{>})$. This is because that it is possible that $e \not\sqsubseteq \textit{MS}(\textit{EPQ}_1^{>})$ because of interaction between actions of multiple priorities. One such example is shown in \figurename~\ref{fig:introduce gap for EPQ1Lar}. We call the time interval from $\textit{ret}(\textit{put},x)$ to $\textit{cal}(rm,x)$, or from $\textit{ret}(\textit{put},x)$ when $\textit{cal}(rm,x)$ does not exist, the interval of item $x$. In \figurename~\ref{fig:introduce gap for EPQ1Lar}, we draw the interval of each item by dashed line. Here we assume that $p_1 <_{\mathbb{P}} p_4$, $p_2 <_{\mathbb{P}} p_4$ and $p_3 <_{\mathbb{P}} p_4$. The reason of $e \not\sqsubseteq \textit{MS}(\textit{PQ}_1^{>})$ is that, each time point from $\textit{cal}(\textit{rm},b)$ to $\textit{ret}(\textit{rm},b)$ is in interval of some item with smaller priority.

\begin{figure}[htbp]
  \centering
  \includegraphics[width=0.4 \textwidth]{figures/PIC-HIS-INTRO-GAP-EPQ1L.pdf}
%\vspace{-10pt}
  \caption{An execution that does not linearizable w.r.t $\textit{MS}(\textit{EPQ}_1^{>})$}
  \label{fig:introduce gap for EPQ1Lar}
\end{figure}

In Appendix \ref{sec:appendix proof and definition in section co-regular of EPQ1Lar}, we introduce the notion of left-right constraint, which is a graph, and the existence of cycle though item with maximal priority in it formally modelled the case in \figurename~\ref{fig:introduce gap for EPQ1Lar}. When there is a cycle $d_1 \rightarrow \ldots \rightarrow d_m \rightarrow x \rightarrow d_1$ through item $x$ with maximal priority in left-right constraint, we say that $x$ is covered by $d_1,\ldots,d_m$ For example, in \figurename~\ref{fig:introduce gap for EPQ1Lar}, $b$ is covered by $f$, $e$ and $d$. We then prove that getting rid of cycle though item with maximal priority in left-right constraint is enough for ensure linearizable w.r.t $\textit{MS}(\textit{EPQ}_1^{>})$, as stated by the following lemma:

@@

Assume that $x$ is covered by $d_1,\ldots,d_m$, we can safely rename $x$ into $b$, rename $d_1,\ldots,d_m$ into $a$ and rename all other item into $c$ by data-independence. Such execution can be recognized by witness automata, since between the first $\textit{ret}(\textit{put},a,\_)$ and the last $\textit{cal}(\textit{rm},a)$ (if exists), $\textit{ret}(\textit{put},a,\_)$ and $\textit{cal}(\textit{rm},a)$ occurs in pair and there is no need for count. In Appendix \ref{sec:appendix proof and definition in section co-regular of EPQ1Lar}, we construct a set $\textit{Auts}_{\textit{1-lar}}$ of witness automata, and use $\textit{Auts}_{\textit{1-lar}}$ to prove that $\textit{EPQ}_1^{>}$ is co-regular, as stated by the following lemma.

@@

In this subsection, we briefly introduce the idea for proving co-regular of $\textit{EPQ}_1^{=}$. The proof of this subsection can be found in Appendix \ref{sec:appendix proof and definition in section co-regular of EPQ1Equal}.

Given a data-differentiated $\_$-execution $e$ such that $\textit{last}(e) = \textit{EPQ}_1^{=}$. Lemma \ref{lemma:automata for extended priority queue with single priority} and $\textit{Auts}_{\textit{1-lar}}$ are sill not enough for ensure that $e \sqsubseteq \textit{MS}(\textit{EPQ}_1^{=})$. This is because that given items $a$ and $b$ with maximal priority, it is possible that all the possible linearization point of $\textit{rm}(b)$ are disabled by $\textit{rm}(b)$. We give an example of such execution $e$ in \figurename~\ref{fig:introduce pb order}. The execution $e$ of \figurename~\ref{fig:introduce pb order} is not linearizable w.r.t $\textit{EPQ}$, even if $h \vert_{p_1}$ and $h \vert_{p_4}$ both does, and either $a$ or $b$ is covered by items with smaller priority. The explanation of $e \not\sqsubseteq \textit{EPQ}$ is as follows: Since $\textit{put}(a,p_4) <_{\textit{hb}} \textit{put}(b,p_4)$, the linearization points of $\textit{rm}(a)$ should before the linearization point of $\textit{rm}(b)$. According to Lemma \ref{lemma:Lin Equals Constraint for EPQ1Lar}, the time intervals identified with dotted lines in \figurename~\ref{fig:introduce pb order} are are possible position to locate the linearization point of $\textit{rm}(b)$. However, each of them is before $\textit{cal}(\textit{rm},a)$.

@@

Let us introduce put-before order $<_{\textit{pb}}$ to formally state that ``an item is putted before another item''. Given a data-differentiated $\_$-execution $e$ and two items $a,b$ with maximal priority, $a <_{\textit{pb}} b$, if one of the following cases holds: (1) $\textit{put}(a,\_) <_{\textit{hb}} \textit{put}(b,\_)$, (2) $\textit{rm}(a) <_{\textit{hb}} \textit{rm}(b)$, or (3) $\textit{rm}(a) <_{\textit{hb}} \textit{put}(b,\_)$. Sometimes we use $a <_{\textit{pb}}^A b$, $a <_{\textit{pb}}^B b$ and $a <_{\textit{pb}}^C b$ to explicitly distinguish above three cases. Let $<_{\textit{pb}}^*$ be the transitive closure of $<_{\textit{pb}}$. Intuitively, $a <_{\textit{pb}}^* b$ means that we can infer that $a$ should be putted earlier than $b$ with the help of several other items.

@@

\begin{definition}\label{def:gap-point for matched put and rm operations}
Given a data-differentiated $\textit{pri}_x$-execution $e$ and two method events $\textit{put}(x,\textit{pri}_x),\textit{rm}(x)$ of $e$. We say that a time-point $o$ is a gap-point of $x$, if $o$ is after $\textit{call}(\textit{put},x,\textit{pri}_x)$ and $\textit{call}(\textit{rm},x)$, is before $\textit{ret}(\textit{rm},x)$, and is not in interval of any items with smaller priority.
\end{definition}

The case of \figurename~\ref{fig:introduce pb order} can be formally described as follows: $a <_{\textit{pb}}^* b$, while the right-most gap-point of $b$ is before $\textit{cal}(\textit{rm},a)$ or $\textit{cal}(\textit{put},a,p_4)$. The following lemma states that getting rid of such case is enough for ensure linearizable w.r.t $\textit{MS}(\textit{PQ}_1^{=})$.

\begin{restatable}{lemma}{EPQOneEqualAsPBandGP}
\label{lemma:EPQ1Equal as pb order and gap-point}
Given a data-differentiated $\textit{pri}$-execution $e$ with $\textit{last}(e) = \textit{EPQ}_1^{=}$. $e \not\sqsubseteq \textit{MS}(\textit{EPQ}_1^{=})$, if and only if there exists $x$ and $y$ with maximal priority $\textit{pri}$ in $e$, such that $y <_{\textit{pb}}^* x$, and the rightmost gap-point of $x$ is before $\textit{cal}(\textit{put},y,\textit{pri})$ or $\textit{cal}(\textit{rm},y)$ in $e$.
\end{restatable}

@@

We have already intuitively explain the proof of the $\textit{if}$ direction. To prove the $\textit{only if}$ direction, we prove its contrapositive. %, where we already know that $\forall x, y$ with maximal priority, if $y <_{\textit{pb}}^* x$, then the rightmost gap-point of $x$ is after $\textit{cal}(\textit{put},y,\textit{pri})$ and $\textit{cal}(\textit{rm},x)$. We need to prove that $e \sqsubseteq \textit{MS}(\textit{EPQ}_1^{=})$. %Or we can say, we need to explicitly construct linearization of $e$.
